What Hast Earth to do With Mars: An explanation of Urbit's IPC interfaces

### 2024-02-23 mss.tex
added abstract and conclusion

\begin{abstract}
In this groundbreaking paper, we embark on a cosmic journey with the renowned character Ernest P. Worrell as he ventures into the unexplored realm of Martian computing. Drawing inspiration from Ernest's comically ingenious encounters with everyday challenges, we investigate the foundations of what we term "Artificial Stupidity." As Ernest grapples with Martian technology, we delve into the intricacies of programming errors, algorithmic missteps, and the curious phenomena that arise when human-like intelligence meets extraterrestrial computing systems. Our analysis sheds light on the unexpected intersections between humor, artificial intelligence, and the cosmic absurdity of Martian software. Join us in this interplanetary exploration as we unravel the mysteries of Artificial Stupidity through the lens of Ernest's interstellar escapades.
\end{abstract}

@@

\section{Introduction}
%Section 1: can do explicit callback to Moron Labs post (we should add that to the standard mss.bib)
% When did urbit backoff on the earth only calls mars? (eyre and iris) Urbit had to run on a VM
In the realm of software, two worlds exist in stark contrast: 
Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
This article sets forth on an expedition to unravel the mysteries of this Martian software landscape, 
exploring how the intricate processes of terrestrial technology can establish a dialogue with the Martian code.

Unlike Earth, where software is a tapestry of evolution and patchwork, Mars presents a realm of purity. 
It is a world where software exists as a Maxwellian construct, untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments. 
Yet, in this isolation lies a profound awareness - an understanding by the Martians that there exist diverse forms of computing with which they must engage. 
Mars, in its wisdom, accommodates these external entities, allowing them to interact in a dialect familiar to its inhabitants: the language of nouns.

In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says ``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth. The latter would be quite impossible, since Earth code is non-Maxwellian. There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.'' This however has never been true for Urbit as it was implemented. Urbit has two vanes \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and curl, respectively). One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.

Our journey into this uncharted territory begins with a deep dive into the architecture of Urbit, focusing on the foundational layer upon which Arvo, Urbit's kernel, operates.
This exploration will illuminate the operational dynamics and philosophical underpinnings that make Urbit a universe apart. 
Following this, we delve into the intricacies of two pivotal components of Urbit’s ecosystem: 
\texttt{\%khan} and \texttt{\%lick}. 
These vanes serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of control and collaboration with Arvo.
Through this exploration, we aim to demystify the enigma of Urbit and bridge the cosmic gap between Earth’s established software paradigms and the Martian vision of computational purity and simplicity.

@@

Arvo's unique architecture avoids what is referred to as "event spaghetti" by maintaining a clear causal chain for every computation. Each chain begins with a Unix I/O event and progresses through a series of steps until the computation's terminal cause. This deterministic nature allows for a high level of predictability and control, a stark contrast to the non-deterministic nature of most Earth-based operating systems.

The kernel's design as a "purely functional operating system" – or more accurately, an "operating function" – underscores its uniqueness. Arvo operates on the principle that the current state is a pure function of its event log, a record of every action ever performed. This determinism is a significant deviation from the norm, where operating systems allow for programmatic alteration of global variables affecting other programs.

Arvo handles nondeterminism in an innovative way. The system, in essence, behaves like a stateful packet transceiver – events are processed, but there's no guarantee of completion, mirroring the behavior of packet dropping in networking. This approach to handling nondeterminism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.

Arvo's determinism is a key feature, stacking it atop a frozen instruction set known as Nock. This concept, though new to operating systems, is not foreign to computing. Similar to how CPU instruction sets like x86-64 are frozen at the chip level, Arvo freezes the instruction set at a higher level, enabling deterministic computation. This high-level determinism contrasts sharply with the non-determinism typically found in Earth's operating systems.

Handling nondeterminism in Arvo is akin to a heuristic decision-making process similar to dropping a packet in networking. This approach views Arvo as a stateful packet transceiver, where the completion of events is not guaranteed, a stark difference from traditional computing models. Additionally, because Arvo runs on a VM, it can obtain nondeterministic information, such as stack traces from infinite loops, through the interpreter beneath it, further enhancing its operational capabilities.

The vision of Urbit, with Arvo at its core, is to transition from developer-hosted web services on multiple foreign servers to self-hosted applications on a personal server. Each architectural decision in Arvo aligns with this vision, emphasizing user ownership and management of data. This approach sets it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.

@@

The architecture of Vere is bifurcated into two distinct parts: the \texttt{king} and the \texttt{serf}. The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth. In contrast, the \texttt{serf} operates as the process running the Arvo virtual machine (VM). This division of responsibilities allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.

@@

A core opcode in Urbit's assembly language (\texttt{nock}) is the \texttt{hint} opcode (opcode 11) this allows Urbit to pass hints to Vere which produces effects on Earth. This bridges the Earth/Mars gap and allows for vanes to interact outside of Mars.

Key to Vere's functionality is how it processes events. Events from the Unix environment are passed to Vere, which then translates and injects them into Arvo. This mechanism demonstrates a profound integration between Vere and Arvo, where Vere possesses direct knowledge of Arvo's state. Notably, Vere can 'scry' or query Arvo's state without altering it, retrieving information as needed. This capability allows Vere to maintain the integrity of Arvo's deterministic model while enabling dynamic interaction with the outside world.

Urbit's use of Unix as its BIOS is a pivotal aspect of its runtime operation. By leveraging the robust, widely-used Unix system, Vere ensures that Urbit can run on a broad range of hardware platforms, effectively making Urbit's Martian technology accessible and operable in Earth's diverse computing environments. This strategic use of Unix not only provides the necessary I/O and optimizations for Arvo but also ensures that Urbit's advanced and unique software architecture can function in tandem with the existing, established hardware and software ecosystems of Earth.

@@

Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem. It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad. This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.

Khan allows threads to be triggered from outside of Urbit. A thread is a monadic function that takes arguments and produces a result. It may perform input and output while running, so it is not a pure function. A thread's strength is that it can easily perform complex IO operations. It uses what's often called the IO monad (plus the exception monad) to provide a natural framework for IO. A thread's weakness is that it's impermanent and may fail unexpectedly. In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails. When code is upgraded, it's impossible to upgrade a running thread, so it fails.

Khan's role in Urbit can be likened to that of a control plane. Its primary function is to execute threads via a Unix Socket and relay the results back. This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.

At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results. These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling. The exception monad further ensures robust management of potential failures and unexpected events.

Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server. Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.

@@

\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. It is a part of the "King" (a.k.a. "Urth") process. It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.

The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane. It serves as a way for unix processes to receive insights about what is happening on urbit. It also allows for pass commands either \%khan and the injection of raw kernel moves into the system. The implication of this means that whoever is running your urbit runtime controls your urbit.

From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}. 
The newt-encoded format is:

@@

\begin{itemize}
  \item \texttt{V} is the version
  \item \texttt{B} is the jam size in bytes (little endian)
  \item \texttt{J} is the jammed noun (little endian)
\end{itemize}

This structure allows for a variety of commands to be executed, including

\begin{enumerate}
  \item \texttt{\%ovum} - the injection of raw kernel moves 
  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
  \item \texttt{\%peek} - namespace scry requests into Arvo
  \item \texttt{\%peel} - emulated namespace scry requests into Vere
\end{enumerate}

A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
\begin{itemize}
  \item \texttt{request-id} matches the input \texttt{request-id}
  \item \texttt{output} depends on the \texttt{command}
\end{itemize}

@@

Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.  (This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.

\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports. Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.  These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
The connexions are made via filepaths in \texttt{.urb/dev} of the pier.

The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.

@@

The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.  The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.

@@

Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.  See e.g. \texttt{\%++ap-generic-take}.  This lets multiple agents share sockets with the same name, and each agent can have its own folder.

@@

As Urbit had to deal with the reality of running on Earth software, Developers had to come up with various ways of communicating between these systems. Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic.

Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the original idea of communication between Earth and Mars. These vanes force Earth to speak to Mars in a language it can understand, nouns.

>>>>

\begin{abstract}
In the evolving landscape of digital computation, the interaction between conventional software paradigms and innovative systems like Urbit presents a unique challenge.
This paper delves into the architectural intricacies and interfaces of Urbit, contrasting its pristine, Maxwellian principles against the complex, evolved nature of Earth's software ecosystems.
Through a comprehensive examination of Urbit’s kernel (Arvo) and runtime (Vere), alongside its high-level and low-level IPC interfaces (\texttt{\%khan} and \texttt{\%lick}, respectively), we elucidate the mechanisms that facilitate communication and control between these disparate computing realms. 
Our analysis aims to demystify Urbit's architecture and underscore its potential to bridge the gap between the familiar digital environments of Earth and the untapped possibilities of Martian computing.

%  This paper explores the innovative approaches undertaken by developers to enable effective communication between the Urbit system (Mars) and conventional software systems (Earth). Initially, Urbit's integration with external systems relied on the use of two specific vanes, \texttt{\%eyre} and \texttt{\%iris}, serving as HTTP client and server respectively. This setup, however, presented challenges to the ideal of autonomous operation envisaged for Urbit, traditionally characterized by the principle that ``Earth always calls Mar; Mars never calls Earth.''

%  Addressing these challenges, the paper introduces the development of two new vanes, \texttt{\%khan} and \texttt{\%lick}, which embody the core principle of uncontaminated communication. These advancements allow for a refined interaction model where Earth-based systems communicate with Urbit in a Urbit's own language, nouns.

%The implications of these developments are significant, offering a novel framework for the integration of distinct computational ecosystems. This paper discusses the technical and conceptual underpinnings of these vanes, their development process, and the broader implications for Urbit's interaction with conventional software systems.

\end{abstract}

@@

\section{Introduction}

In the realm of software, two worlds exist in stark contrast: 
Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
This article will explore how Earth software establish a communication with the Martian code.

Unlike Earth, where software is a tapestry of evolution and patchwork, Mars presents a realm of purity. 
It is a world where software exists as a Maxwellian construct, it is defined by a small set of axioms \citep{Nock4k}. Urbit code is untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments. 
Yet, in this isolation lies a profound awareness - an understanding by the Martians that there exist diverse forms of computing with which they must engage. 
Mars, accommodates these external entities by allowing them to interact in the Martian language of nouns.

In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says 
``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth. The latter would be quite impossible, since Earth code is non-Maxwellian. There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.''
This however has never been true for Urbit as it was implemented. Urbit has two vanes (best thought of as kernel modules) \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and client, respectively). One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.

This paper begins with a brief explanation of Urbit's architecture, focusing on Arvo (Urbit's kernel) and Vere (Urbit's Runtime). 
Following this, is an explanation of three core parts of the Urbit OS: 
\texttt{\%khan}, \texttt{conn.c} and \texttt{\%lick}. 
These serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of command and control of the entire system.
Through this exploration, we aim to demystify the architecture of Urbit and bridge the gap between Earth’s established software paradigms and the Martian vision of computation.

@@

Arvo's unique architecture avoids what is referred to as ``event spaghetti'' by maintaining a clear causal chain for every computation. 
Each chain begins with a Unix I/O event and progresses through a series of steps until the computation either completes successfully or fails.
This deterministic nature allows for a high level of predictability and control, a stark contrast to most Earth-based operating systems.

The kernel's design as a ``purely functional operating system'' – or more accurately, an ``operating function''" – underscores its uniqueness.
Arvo operates on the principle that the current state is a pure function of its event log, a record of every action ever performed.
This determinism is a significant deviation from the norm, where operating systems allow for programmatic alteration of global variables affecting other programs.

Arvo handles non-determinism in an innovative way. 
The system, in essence, behaves like a stateful packet transceiver -- events are processed, but there's no guarantee of the event successfully completing.
This mirrors the behavior of packet dropping in networking.
This approach to handling non-determinism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.
Additionally, because Arvo runs on a VM (Vere), it can obtain non-deterministic information, such as stack traces from infinite loops, through the interpreter beneath it injecting events.

@@

The architecture of Vere is split into two distinct parts: the \texttt{king} and the \texttt{serf}.
The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth.
In contrast, the \texttt{serf} operates as the process running the Arvo virtual machine (VM).
This division of responsibilities allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.

@@

A core opcode in Urbit's assembly language (Nock) is the \texttt{hint} opcode (opcode 11). 
The \texttt{hint} opcode passes some value to the interpreter. 
The interpreter can then choose what higher order action to take based on this value.
For example, the simplest use of the \texttt{hint} opcode is the ordinary ``debug printf''.
A more complex use of the opcode allows Urbit to pass hints to Vere which produces effects on Earth (send a network packet, get a keyboard input, etc.); bridging the Earth/Mars gap and allowing for vanes to interact outside of Mars.

Key to Vere's functionality is how it processes events. 
Events from the Unix environment are passed to Vere, which then translates and injects them into Arvo.
This mechanism demonstrates a profound integration between Vere and Arvo, where Vere possesses direct knowledge of Arvo's state. 
Notably, Vere can ``scry'' or query Arvo's state without altering it, retrieving information as needed. 
This capability allows Vere to maintain the integrity of Arvo's deterministic model while enabling dynamic interaction with the outside world.

Urbit's use of Unix as its BIOS is an important aspect of its runtime operation. 
By leveraging the robust, widely-used Unix system, Vere ensures that Urbit can run on a broad range of hardware platforms, effectively making Urbit's Martian technology accessible and operable in Earth's diverse computing environments. 
This strategic use of Unix not only provides the necessary I/O and optimizations for Arvo but also ensures that Urbit's advanced and unique software architecture can function in tandem with the existing, established hardware and software ecosystems of Earth.

@@

\texttt{\%khan}, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
It is designed to efficiently handle complex I/O operations, employing an I/O monad coupled with an exception monad.
This approach allows \texttt{\%khan} to adeptly manage complex I/O tasks while effectively handling potential failures.

At its core, \texttt{\%khan}'s design revolves around the use of `threads', which are monadic functions designed to handle arguments and produce results.
These threads are optimized for I/O operations, leveraging the I/O monad for structured and efficient task handling. 
The exception monad further ensures robust management of potential failures and unexpected events.
The weakness of threads is that it's impermanent and may fail unexpectedly. 
Since it is usually relying on some external 
In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails.
When code is upgraded, it's impossible to upgrade a running thread, so it fails.

\texttt{\%khan} was conceived as a way to control Urbit ships from the exterior using threads.
The concept evolved a fair bit from proposal to implementation. 
In practice, \texttt{\%khan} is essentially an interface wrapper for Spider-based threads, which produces a somewhat strange (but not unprecedented) situation in which a vane relies on a piece of userspace infrastructure to function correctly.
\texttt{\%spider} is an agent for transient thread-level operations.

\texttt{\%khan} allows for pre-written threads that allow for easy hosting and maintenance to be bundled and distributed; helping hosting companies and self-hosted ships to easily run efficiently.

@@

\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. 
It is a part of the ``King'' process. 
It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.

The functionality of \texttt{conn.c}, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of Urbit's control plane.
It serves as a way for Unix processes to receive insights about what is happening on Urbit. 
It also allows for pass commands either \texttt{\%khan} and the injection of raw kernel moves (events) into the system.

From a technical perspective, \texttt{conn.c} accepts newt-encoded \texttt{++jam}ed nouns (defined below) which take the form \texttt{[request-id command arguments]}. 
The newt-encoded format is:

@@

Where 
\begin{itemize}
  \item \texttt{V} is the version (currently 0)
  \item \texttt{B} is the size of the \texttt{++jam}ed noun in bytes (little endian)
  \item \texttt{J} is the \texttt{++jam}ed noun (little endian)
\end{itemize}

This structure allows for a variety of commands to be executed, including:

\begin{enumerate}
  \item \texttt{\%ovum} - the injection of raw kernel moves 
  \item \texttt{\%fyrd} - a direct shortcut to \texttt{\%khan} commands
  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
  \item \texttt{\%peek} - namespace scry requests into Arvo
  \item \texttt{\%peel} - emulated namespace scry requests into Vere
\end{enumerate}

A valid \texttt{conn.c} command produces a newt-encoded \texttt{++jam}ed noun with type \texttt{[request-id output]}, where:
\begin{itemize}
  \item \texttt{request-id} matches the input \texttt{request-id}
  \item \texttt{output} depends on the \texttt{command}
\end{itemize}

@@

Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different use-case than \texttt{\%khan}: to allow external processes, in particular hardware drivers, to intercommunicate with Urbit. 
\texttt{\%lick} focuses on creating a generic noun interface over domain sockets.

\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports.
Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
\texttt{\%lick} works by opening a Unix socket for a particular process, which allows serialized IPC communications.  These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
The IPC ports texttt{\%lick} creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
The connections are made via file path in \texttt{.urb/dev} of the pier.

The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the \texttt{++jam}ed noun, check the mark and (most likely) convert the noun into a native data structure.

@@

The vane definition of \texttt{\%lick} is even simpler than \texttt{\%khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.

@@

Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly (see e.g. \texttt{\%++ap-generic-take}). 
This lets multiple agents share sockets with the same name, while placing an agents sockets in a unique file path.

@@

This exploration into Urbit's IPC interfaces, specifically \texttt{\%khan} and \texttt{\%lick}, has illuminated the profound architectural innovations that distinguish Urbit from traditional computing paradigms. 
By delving into the inner workings of Arvo and Vere, as well as the mechanisms facilitating communication between these components, we've uncovered a realm where simplicity and determinism reign supreme. 
This study not only demystifies the complex underpinnings of Urbit but also highlights its potential to redefine our interaction with digital ecosystems.

Urbit's design philosophy, rooted in the principles of a more principled and coherent digital universe, challenges us to reconsider the foundations upon which our current software ecosystems are built. 
The contrast between the evolved, patchwork nature of Earth's software and the Maxwellian purity of Urbit's architecture serves as a poignant reminder of the potential for innovation in software design. 
As we stand at the crossroads of Earth's complexity and Mars' simplicity, the lessons drawn from Urbit's IPC interfaces beckon us towards a future where software is not only a tool but a reflection of a more streamlined and purposeful digital existence.

The journey through Urbit's landscape offers a vision of what digital computation could become: a domain where efficiency, clarity, and integrity are not aspirational but foundational. 
As we continue to navigate the evolving landscape of software development, the insights garnered from this analysis of Urbit will undoubtedly inspire further exploration and innovation. 
In embracing the principles elucidated by Urbit's design, we pave the way for a future where the digital and human experiences are more intimately aligned, fostering a computing environment that is both revolutionary and deeply resonant with the principles of simplicity and purpose that guide our exploration of the digital frontier.

%As Urbit had to deal with the reality of running on Earth software, developers had to come up with various ways of communicating between these systems.
%Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic. 
%Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the necessity idea of communication between Earth and Mars. These vanes force Earth to speak to Mars in a language it can understand, nouns.
